\section{The Compliance Crossover Analysis}
\label{sec:crossover}

We compare the TCO model from Section~\ref{sec:tco} against the
expected cost of regulatory non-compliance. The \emph{compliance
crossover point} $N^*$ is the annual decision volume at which the
full-stack accountability infrastructure becomes strictly less expensive
than the risk-weighted expected penalty for an equivalent non-transparent
system.

\subsection{Expected Penalty Model}
\label{subsec:penalty-model}

Let $P_{\text{enf}}$ denote the annual probability that a non-compliant
operator receives an enforcement action, and $E[\text{fine}]$ the
expected monetary penalty conditional on enforcement. We model the
\emph{risk density} $\rho$ as the expected penalty attributable per
decision, computed \emph{per controller} rather than per market:

\[
  \rho = P_{\text{enf}} \times \frac{E[\text{fine}]}{\bar{N}_{\text{controller}}}
\]

\noindent where $\bar{N}_{\text{controller}}$ is the average annual
decision volume of a fined controller. This formulation correctly
attributes enforcement risk to the scale of the individual operator,
not to aggregate market statistics.

The expected annual penalty cost for an operator with $N$ decisions is:

\[
  C_{\text{penalty}}(N) = \rho \times N.
\]

\subsection{Parameter Estimation by Regulatory Regime}
\label{subsec:params}

\paragraph{GDPR (Art.~22 automated decision-making).}
Based on DLA~Piper enforcement data~\cite{gdpr-fines-2025}, we estimate
$P_{\text{enf}} \approx 1\%$ per year for controllers operating
automated decision-making at scale. The median fine for Art.~22
violations by mid-to-large controllers is approximately
\euro{}10M~$\approx$ \$10.8M, and such controllers typically process
$\bar{N}_{\text{controller}} \approx 10^6$ decisions/year. Therefore:
\[
  \rho_{\text{GDPR}} = 0.01 \times \frac{\$10.8\text{M}}{10^6}
  = \$0.108 \times 10^{-1} = \$\mathbf{0.01}\text{/decision}.
\]

\paragraph{EU AI Act (High-Risk, Art.~99).}
With enforcement projected to begin at scale from 2026, we apply
a conservative $P_{\text{enf}} \approx 2\%$/year for high-risk AI
systems. Fines of up to 3\% of global turnover apply; for a
mid-market operator with \$100M revenue this implies \$3M. With
$\bar{N}_{\text{controller}} \approx 3 \times 10^6$ decisions/year:
\[
  \rho_{\text{AI Act}} = 0.02 \times \frac{\$3\text{M}}{3 \times 10^6}
  = \$\mathbf{0.02}\text{/decision}.
\]

\paragraph{SEC off-channel communications (regulated firms).}
SEC enforcement history (2021--2024) shows $P_{\text{enf}} \approx 5\%$
for broker-dealers, with fines averaging \$125M. Regulated firms
typically process $\bar{N}_{\text{controller}} \approx 6 \times 10^6$
communication-linked decisions/year:
\[
  \rho_{\text{SEC}} = 0.05 \times \frac{\$125\text{M}}{6 \times 10^6}
  \approx \$\mathbf{1.04}\text{/decision}.
\]

\subsection{Computing the Crossover Point}
\label{subsec:crossover-calc}

Since variable TCO costs are negligible (Section~\ref{sec:tco}), we
approximate $\text{TCO}(N) \approx C_{\text{fixed}} = \$5{,}000$/year
for $N < 10^7$. The crossover condition becomes:

\[
  N^* = \frac{C_{\text{fixed}}}{\rho}
\]

Table~\ref{tab:crossover} reports $N^*$ per regulatory regime.

\begin{table}[h]
\centering
\caption{Compliance Crossover Points by Regulatory Regime}
\label{tab:crossover}
\begin{tabular}{lrrr}
\toprule
\textbf{Regime} & $P_{\text{enf}}$ & $\rho$ & $N^*$ \\
\midrule
GDPR Art.~22           & 1\%  & \$0.010  & 500{,}000 \\
EU AI Act (high-risk)  & 2\%  & \$0.020  & 250{,}000 \\
SEC off-channel        & 5\%  & \$1.040  & 4{,}808   \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Multi-Regime Operators and the Reported Threshold}
\label{subsec:multiregime}

Many operators are subject to multiple regulatory regimes
simultaneously. A Brazilian financial institution processing EU
citizens' credit data faces GDPR, AI~Act, and LGPD exposure
concurrently. For such operators, the effective risk density is the
sum across regimes:
\[
  \rho_{\text{combined}} = \rho_{\text{GDPR}} + \rho_{\text{AI Act}}
  = \$0.01 + \$0.02 = \$0.03\text{/decision}.
\]

The corresponding crossover is:
\[
  N^*_{\text{combined}} = \frac{\$5{,}000}{\$0.03} \approx 167{,}000
  \text{ decisions/year.}
\]

However, the GDPR single-regime crossover at 500{,}000 decisions/year
represents the most conservative bound applicable to operators subject
only to GDPR --- the regime with the lowest historical enforcement
frequency among those analysed. We adopt this as the reported threshold
in the abstract to avoid overstating the economic case; all claims
derived from the analysis hold at this bound.

\subsection{Discussion}
\label{subsec:crossover-discussion}

The analysis yields three actionable results.

First, for SEC-regulated financial firms, $N^* \approx 5{,}000$
decisions/year --- a threshold almost certainly surpassed by any
operational trading or compliance system. For this class of operator,
deploying the accountability stack is rational even at very small scale.

Second, for operators subject solely to GDPR, the crossover occurs at
$N^* = 500{,}000$ decisions/year. This is easily exceeded by national
credit-scoring, healthcare triage, and hiring systems --- precisely the
high-risk AI deployments targeted by both GDPR Art.~22 and the EU~AI~Act.

Third, the compliance-credit mechanism proposed in
Section~\ref{sec:credits} shifts the effective crossover point further
downward by converting transparent behaviour into a positive regulatory
asset, reducing $P_{\text{enf}}$ for certified operators and thereby
raising the post-credit $N^*$ for non-certified ones.
