\section{Total Cost of Ownership (TCO) Model}
\label{sec:tco}

We define the cost model for operating the BTV accountability stack
(Papers~1--3~\cite{btv2026,btv-t-repo,btv-ar2026}) at production scale.
The TCO comprises fixed infrastructure costs and variable compute/storage
costs. We analyse three representative deployment tiers: Startup
($10^4$ decisions/year), Enterprise ($10^6$ decisions/year), and
Hyperscale ($10^8$ decisions/year).

\subsection{Fixed Infrastructure Costs}
\label{subsec:fixed}

The transparency stack requires two dedicated components not present in
standard AI pipelines:

\begin{enumerate}
  \item \textbf{Transparency Log Server.} A high-availability
  append-only log (e.g., Trillian or equivalent) running on managed
  cloud infrastructure with replication across two availability zones.
  Estimated at \$2{,}500/year on AWS~EC2/RDS.

  \item \textbf{Hardware Security Module (HSM).} Required to protect
  the Ed25519 Log Authority signing key (Paper~2) and the Pedersen
  blinding factors for redaction keys (Paper~3). Managed HSM
  (e.g., AWS CloudHSM) costs approximately \$2{,}500/year.
\end{enumerate}

\noindent Total fixed cost $C_{\text{fixed}} \approx \$5{,}000$/year,
independent of decision volume.

\subsection{Variable Compute Costs}
\label{subsec:compute}

Compute costs are derived from the benchmarks in Papers~2 and~3.
We assume a cloud compute unit cost of \$0.05 per vCPU-hour.

\begin{itemize}
  \item \textbf{Paper~1 (Compile-time binding).} Zero marginal runtime
  cost. One-time engineering integration estimated at $\sim$40~hours.

  \item \textbf{Paper~2 (Transparency log append).} Latency
  4.73~ms/decision, throughput $\approx$760~decisions/s per vCPU.
  At $10^8$ decisions/year ($\approx$3.2 decisions/s average),
  sustained load requires $< 0.01$ vCPU. Annual compute cost
  $< \$2$/year at any modelled scale.

  \item \textbf{Paper~3 (ZK-SNARK proving for redaction).} Assuming a
  GDPR-typical redaction request rate of 0.1\%, each $10^6$-decision
  deployment generates $\sim$1{,}000 redaction batches/year. At
  $\approx$2.1~s/batch on a Xeon core, total proving time is
  $\approx$35 core-minutes/year; cost $\approx$\$0.03/year.
\end{itemize}

\subsection{Variable Storage Costs}
\label{subsec:storage}

Each decision produces two persistent artefacts: a \texttt{DeliveryToken}
(56~bytes, Paper~2) and a \texttt{InclusionReceipt} (264~bytes, Paper~2).
Total footprint: 320~bytes/decision. At cloud object-storage rates
(\$0.023/GB/month, \$0.276/GB/year):

\[
  C_{\text{storage}}(N) = N \times 320\,\text{B} \times \$0.276/\text{GB}
  = N \times \$8.8 \times 10^{-8}.
\]

At $10^8$ decisions/year, storage cost $\approx$\$8.80/year --- less than
one hourly instance cost.

\subsection{TCO Summary}
\label{subsec:tco-summary}

Table~\ref{tab:tco} summarises the annualised TCO across the three
scales. Variable costs are negligible relative to fixed infrastructure;
the cost per decision is dominated entirely by amortisation of
$C_{\text{fixed}}$.

\begin{table}[h]
\centering
\caption{Annualised TCO of the BTV Accountability Stack}
\label{tab:tco}
\begin{tabular}{lrrr}
\toprule
\textbf{Cost component}
  & \textbf{Startup} ($10^4$)
  & \textbf{Enterprise} ($10^6$)
  & \textbf{Hyperscale} ($10^8$) \\
\midrule
Fixed infrastructure        & \$5{,}000    & \$5{,}000    & \$5{,}000    \\
Compute (P2 append)         & $<$\$1       & $<$\$1       & $<$\$2       \\
Compute (P3 ZK proving)     & $<$\$0.01    & $<$\$0.03    & $<$\$3       \\
Storage (P2 receipts)       & $<$\$0.01    & $<$\$0.09    & $<$\$9       \\
\midrule
\textbf{Total/year}         & \textbf{\$5{,}000} & \textbf{\$5{,}010} & \textbf{\$5{,}014} \\
\textbf{Per-decision cost}  & \textbf{\$0.50}    & \textbf{\$0.005}   & \textbf{\$0.00005} \\
\bottomrule
\end{tabular}
\end{table}

\noindent The knee of the cost curve occurs near $N \approx 50{,}000$
decisions/year, below which the fixed infrastructure cost produces a
per-decision cost exceeding \$0.10. Above this threshold, the marginal
cost of verifiable accountability is effectively zero: all additional
decisions are verified at storage cost alone.
