\section{The Compliance-Credit Mechanism}
\label{sec:credits}

The crossover analysis in Section~\ref{sec:crossover} treats
non-compliance risk as a fixed external cost that is either borne
(opaque strategy) or avoided (transparent strategy). This framing
omits a second-order effect present in all major regulatory regimes:
operators that can demonstrate structural compliance controls at
enforcement time receive materially reduced penalties. We formalise
this as a \emph{compliance credit} and show that it changes both the
quantitative crossover threshold and the qualitative incentive
structure.

\subsection{Regulatory Foundations}
\label{subsec:legal-foundations}

Mitigating factors for demonstrated compliance are explicit in the
relevant frameworks:

\begin{itemize}
  \item \textbf{GDPR Art.~83(2)(d),(k):} Fines are modulated by
  ``the degree of cooperation with the supervisory authority'' and
  ``any other aggravating or mitigating factor.'' Implementing
  Privacy-by-Design measures (Art.~25) is an established mitigating
  factor in supervisory practice.

  \item \textbf{GDPR Art.~42:} Certification schemes approved by a
  national supervisory authority ``may be used as an element by
  which to demonstrate compliance with the requirements'' --- creating
  a direct path for BTV-stack certification to constitute
  enforceable evidence of due diligence.

  \item \textbf{AI Act Art.~69:} ``Codes of conduct'' adopted by
  operators ``may contribute to demonstrating compliance'' with
  high-risk AI requirements, directly enabling safe-harbour
  treatment for certified deployments.

  \item \textbf{US Sentencing Guidelines \S8B2.1:} Organisations
  with an effective compliance and ethics programme may receive
  up to a 95\% reduction in criminal penalties. The key criterion ---
  that the programme ``prevent and detect'' violations --- is
  satisfied by design in the BTV stack, where violations are
  structurally impossible to conceal rather than merely discouraged
  by policy.
\end{itemize}

\subsection{Formal Definition of the Credit}
\label{subsec:formal-credit}

We introduce a mitigation factor $\delta \in [0, 1]$ representing the
fraction of the expected penalty avoided by an operator who deploys
and maintains the certified BTV accountability stack. The expected
penalty for a transparent operator becomes:

\[
  C_{\text{penalty}}^{\text{transparent}}(N)
  = (1 - \delta) \times \rho \times N.
\]

We estimate $\delta = 0.50$ (50\% reduction) as a conservative
baseline grounded in supervisory practice: GDPR enforcement decisions
consistently apply 30--60\% reductions for demonstrated
Privacy-by-Design implementation~\cite{gdpr-fines-2025}, and the
BTV stack satisfies a stronger standard (structural non-repudiability)
than process-level controls.

\subsection{Corrected Crossover with Credit}
\label{subsec:credit-crossover}

An operator chooses between two strategies:

\begin{enumerate}
  \item \textbf{Opaque:}
  $C_{\text{opaque}}(N) = \rho \times N$
  (no infrastructure cost; full penalty exposure).

  \item \textbf{Transparent:}
  $C_{\text{transparent}}(N) = \text{TCO}(N) + (1-\delta)\rho N$
  (infrastructure cost plus residual penalty exposure).
\end{enumerate}

The transparent strategy is preferred when
$C_{\text{transparent}} < C_{\text{opaque}}$, i.e.,

\[
  \text{TCO}(N) + (1-\delta)\rho N < \rho N
  \implies \text{TCO}(N) < \delta \rho N.
\]

Approximating $\text{TCO}(N) \approx C_{\text{fixed}}$ (valid for
$N < 10^7$, cf.\ Section~\ref{sec:tco}), the crossover point with
credit is:

\[
  \boxed{N^*_{\text{credit}} = \frac{C_{\text{fixed}}}{\delta \, \rho}.}
\]

Note that $N^*_{\text{credit}} = N^*_{\text{no credit}} / \delta$:
the credit \emph{raises} the no-credit crossover threshold because it
partially subsidises the opaque strategy by reducing its penalty
exposure. The correct interpretation is not that the credit makes
transparency cheaper in absolute terms, but that it reframes the
\emph{benefit} of the transparent strategy as penalty avoidance
\emph{on top of} penalty reduction:

\[
  \text{Net benefit}(N) = \delta \rho N - C_{\text{fixed}}.
\]

This is positive for $N > N^*_{\text{credit}}$ and grows linearly
with $N$ at rate $\delta \rho$.

\subsection{Crossover Table with Credit}
\label{subsec:credit-table}

Table~\ref{tab:credit-crossover} reports $N^*_{\text{credit}}$ for
$\delta = 0.50$ across the three regulatory regimes established in
Section~\ref{sec:crossover}.

\begin{table}[h]
\centering
\caption{Compliance-Credit Crossover Points ($\delta = 0.50$)}
\label{tab:credit-crossover}
\begin{tabular}{lrrr}
\toprule
\textbf{Regime} & $\rho$ & $N^*$ (no credit) & $N^*$ (with credit) \\
\midrule
GDPR Art.~22          & \$0.010 & 500{,}000   & 1{,}000{,}000 \\
EU AI Act (high-risk) & \$0.020 & 250{,}000   & 500{,}000     \\
SEC off-channel       & \$1.040 & 4{,}808     & 9{,}615       \\
Multi-regime (GDPR + AI Act) & \$0.030 & 167{,}000 & 333{,}000 \\
\bottomrule
\end{tabular}
\end{table}

Under the EU~AI~Act --- the regulatory framework most directly
targeting high-risk AI systems, and the one that entered into
enforcement for high-risk categories in August 2026 --- the
compliance-credit crossover occurs at \textbf{500,000
decisions/year}. This is the threshold reported in the abstract and
corresponds to the practically relevant regime for the deployments
(credit scoring, hiring, healthcare triage) that the BTV stack is
designed to protect.

\subsection{Market Structure: The Carbon-Credit Analogy}
\label{subsec:carbon-analogy}

The compliance-credit mechanism is structurally analogous to the EU
Emissions Trading System (ETS)~\cite{ellerman2010}: operators that
exceed compliance thresholds must purchase credits or face penalties;
those with certified infrastructure generate surplus credits. Three
properties transfer directly:

\begin{enumerate}
  \item \textbf{Certification as the unit of account.} Just as a
  verified tonne of $\text{CO}_2$ is the ETS unit, a
  \emph{verified decision batch} attested by the BTV stack's
  cryptographic receipt is the accountability unit.
  Certification is non-gameable: inclusion proofs cannot be
  fabricated without breaking Poseidon collision resistance
  (Property~3 of Paper~3~\cite{btv-ar2026}).

  \item \textbf{Transferability.} Large operators generating surplus
  compliance credits could transfer them to smaller operators under
  enforcement pressure, creating a compliance-services market and
  reducing the barrier to entry for deployments below the
  $N^*_{\text{credit}}$ threshold.

  \item \textbf{No new legislation required.} GDPR Art.~42
  (certification schemes) and AI Act Art.~69 (codes of conduct)
  already provide the legal mechanism for regulators to issue
  safe-harbour recognition. The compliance credit is an
  administrative instrument, not a legislative one.
\end{enumerate}
