\section{Security Analysis: Resistance to Constitutional Capture}
\label{sec:security}

We analyse the Constitutional Amendment Protocol against a formal
threat model. The primary adversarial goal is \emph{Constitutional
Capture}: modifying the Stone Clauses $\mathcal{L}^*$ in a way that
weakens the power separation guarantees of Paper~5~\cite{btv-cc},
or operating outside the boundaries of a valid mandate.

\subsection{Threat Model}
\label{subsec:threat-model}

We consider a probabilistic polynomial-time adversary $\mathcal{A}$
who may corrupt at most two of the three constitutional branches
(Legislative, Executive, Judicial). Corrupting a branch means
obtaining the signing key of its representative. We assume the
third branch remains honest. We also assume the security of the
underlying signature scheme and ZK proof system (standard
assumptions from Papers 2--3~\cite{btv-t-repo,btv-ar2026}).

\subsection{Attack Vector 1: The Hostile Takeover}
\label{subsec:attack-takeover}

\textbf{Scenario.} A compromised regulator ($\mathcal{L}$) and
compliant operator ($\mathcal{E}$) collude to amend a Stone Clause
$x \in \mathcal{L}^*$---specifically, to remove the requirement for
ZK audit proofs, enabling opaque decision-making.

\begin{enumerate}
  \item $\mathcal{A}$ controls $\sigma_{\mathcal{L}}$ and
  $\sigma_{\mathcal{E}_{\mathsf{rep}}}$.
  \item $\mathcal{A}$ constructs amendment transaction
  $\Delta\mathcal{L}^* = \{\text{remove ZK requirement}\}$.
  \item $\mathcal{A}$ submits $\Delta\mathcal{L}^*$ with both
  signatures.
\end{enumerate}

\textbf{Defence.} By Definition~\ref{def:tripartite}, valid
constitutional amendments require
$\sigma_{\mathcal{L}} \wedge \sigma_{\mathcal{J}} \wedge
\sigma_{\mathcal{E}_{\mathsf{rep}}}$. The adversary lacks
$\sigma_{\mathcal{J}}$. The Monitor's signing key is held by an
accredited independent organisation whose accreditation is maintained
in a public registry (Paper~5, \S5.2~\cite{btv-cc}); the registry
is itself append-only (Paper~2~\cite{btv-t-repo}) and cannot be
modified by $\mathcal{L}$ or $\mathcal{E}$ unilaterally. The
Monitor's economic incentive is to detect violations, not to enable
them: its reputation and accreditation status are public and
audit-dependent. Therefore:
\[
  \Pr[\mathcal{A}\text{ succeeds}] \leq \Pr[\mathcal{A}
  \text{ forges } \sigma_{\mathcal{J}}] \leq \epsilon_{\mathsf{sig}}
\]
where $\epsilon_{\mathsf{sig}}$ is the forgery probability of the
underlying signature scheme (negligible in the security parameter
$\lambda$). The Hostile Takeover fails with overwhelming probability.

\textbf{Residual risk.} If the adversary can also corrupt the
Judicial branch ($k = 3$ corruptions), all three signatures are
available and the amendment succeeds. This is not a failure of the
protocol but a constitutional threshold by design: a unanimous
decision of all three branches to amend a Stone Clause is, in
classical constitutional theory, a legitimate if extreme exercise
of constituent power. The protocol does not prevent unanimous
consent; it makes it detectable and publicly verifiable via the
log, ensuring that society can observe and respond to such an event.

\subsection{Attack Vector 2: The Zombie Institution}
\label{subsec:attack-zombie}

\textbf{Scenario.} The operator continues serving decisions after
the Legislative Mandate expires ($t > t_{\mathsf{exp}}$), claiming
the system operates under the prior mandate by inertia.

\begin{enumerate}
  \item Mandate $M_v = \texttt{Mandate}[\mathcal{L}_v,
  t_{\mathsf{exp}}]$ expires at $t_{\mathsf{exp}}$.
  \item No renewed $M_{v+1}$ has been ratified (\emph{Interregnum}).
  \item $\mathcal{A}$ attempts to serve a new decision $d$ by
  calling \texttt{Verdict::new}.
\end{enumerate}

\textbf{Defence.} By Theorem~\ref{thm:sunset}, \texttt{Verdict::new}
requires a live reference to $M$. At $t > t_{\mathsf{exp}}$, $M$
has been consumed by the expiry check and cannot be borrowed. The
call fails with a Rust borrow error at the \emph{type level}:
there is no runtime path that allows the operator to construct a
valid \texttt{DeliveryToken} without a live mandate. The system
enters Interregnum automatically; no external intervention is
required.

\textbf{Observable signal.} Because the last valid decision in
$\Sigma$ carries timestamp $t_{\mathsf{last}} < t_{\mathsf{exp}}$,
any auditor observing decisions with $t > t_{\mathsf{exp}}$ knows
immediately that the operator is running an uncertified binary---a
binary substitution detectable by hash comparison
(Theorem~\ref{thm:leg}, Paper~5~\cite{btv-cc}). The Zombie
Institution is structurally self-revealing.

\subsection{Summary of Security Properties}
\label{subsec:security-summary}

\begin{table}[h]
\centering
\caption{Security Properties of the Amendment Protocol}
\label{tab:security}
\begin{tabular}{lp{4cm}p{4cm}}
\toprule
\textbf{Attack} & \textbf{Adversarial Capability} & \textbf{Outcome} \\
\midrule
Hostile Takeover & Controls $\mathcal{L}$ and $\mathcal{E}$ (2-of-3) & Amendment rejected; missing $\sigma_{\mathcal{J}}$ \\
Hostile Takeover & Controls all 3 branches & Amendment succeeds but is publicly logged and verifiable \\
Zombie Institution & Runs expired binary & Structurally impossible to produce valid verdicts; self-revealing via hash mismatch \\
Retrospective Capture & Rewrites historical $\mathcal{L}_v$ & Breaks hash chain in $\Sigma$; detected by any auditor \\
\bottomrule
\end{tabular}
\end{table}
