Transparency logs have become a de facto standard for verifiable accountability
in high-stakes AI systems, satisfying auditability requirements under EU AI
Act Art.~86. This immutable architecture conflicts with data protection law
(GDPR Art.~17, LGPD Art.~18), which mandates deletion of personal data upon
request. Existing redactable authenticated data structures address the
mechanics of deletion but fail to prevent the \emph{Selective Deletion Attack}:
a system operator who exploits individual erasure requests to selectively remove
decisions that constitute evidence of algorithmic discrimination, while retaining
favorable entries, thereby distorting the log's statistical history without
triggering any existing integrity alarm.

This paper presents \textbf{Accountable Redaction}, a protocol that enforces
statistical integrity as a first-class verification condition on every log
transition. We model the log as a state machine: a redaction batch $R$
induces a transition $L_t \xrightarrow{R} L_{t+1}$, which is accepted if and
only if
\[
  \text{Valid}\!\left(L_t \xrightarrow{R} L_{t+1}\right) \;\iff\;
  \mathsf{Verify}(\pi_{\mathit{auth}},\, R)
  \;\land\;
  \mathsf{Verify}(\pi_{\mathit{stat}},\, L_t,\, L_{t+1},\, \varepsilon,\, \Pi)
\]
where $\pi_{\mathit{auth}}$ is an authorization proof that each redacted entry
was requested by the data subject or a legitimate authority (preventing
operator-initiated self-serving erasure), and $\pi_{\mathit{stat}}$ is a
ZK-SNARK over Pedersen-committed demographic attributes proving that the
conditional approval rate of every protected group in the fairness policy $\Pi$
shifts by at most~$\varepsilon$ after the batch.

We define $\varepsilon$-statistical consistency formally, prove that the
protocol satisfies four security properties (Redaction Soundness, Content
Indistinguishability, Merkle Continuity, and Operator Non-Frameability), and
give a circuit construction in Noir. The separation of $\pi_{\mathit{auth}}$
from $\pi_{\mathit{stat}}$ is not merely notational: it closes an attack vector
absent from all prior redactable blockchain schemes, in which an operator can
self-authorize erasure requests. Empirical evaluation on a commodity server
gives proof generation times of approximately 2.1~s per batch of 1{,}000
decisions with constant-size proofs ($\approx$3.2~kB), making the protocol
practical for regulated AI deployments. Together with the BTV series
(Papers~1--2), this work closes the full accountability loop: no decision
without evidence, no delivery without persistence, and no redaction without
statistical justification.
