\section{Zero-Knowledge Circuit Construction}
\label{sec:zk}

This section details the implementation of the ZK circuit $\mathcal{C}$
responsible for generating the statistical consistency proof
$\pi_{\mathit{stat}}$. We address the specific challenges of private
demographic aggregation and fixed-point arithmetic verification within the
constraints of Rank-1 Constraint Systems (R1CS).

\subsection{Private Attributes via Pedersen Commitments}
\label{subsec:pedersen}

To satisfy privacy requirements (GDPR Art.~5), demographic attributes are
never exposed in the public Merkle tree leaves. Each leaf $\ell_i$ stores a
Pedersen commitment $C_i = \mathsf{Com}(g_i;\, r_i)$ where $g_i \in
\mathcal{G}$ is the group identifier and $r_i$ is a random blinding factor
chosen at insertion time.

The circuit accesses $g_i$ and $r_i$ as private witnesses and enforces:
\begin{enumerate}
  \item \textbf{Commitment integrity.} The circuit recomputes
  $\mathsf{Com}(g_i;\, r_i)$ using the public generator points and asserts
  equality with the $C_i$ stored in the public leaf. This prevents the
  prover from supplying a fabricated group label to manipulate the
  per-group statistics.
  \item \textbf{Merkle inclusion.} The circuit verifies that the leaf
  containing $C_i$ is reachable via the witness path $\mathit{path}_i$
  from the public root $\mathit{rt}_t$.
\end{enumerate}

\noindent The group label $g_i$ is consumed entirely within the circuit;
the verifier observes only the aggregated outcome (the committed statistics
$\mathsf{Com}(S(L_{t+1}))$), not any per-entry attribute.

\subsection{Fixed-Point Arithmetic via Euclidean Division}
\label{subsec:fixedpoint}

R1CS circuits over prime fields support multiplication and addition but not
native division. To enforce the $\varepsilon$-consistency bound
(Definition~\ref{def:eps-stat}), which requires comparing the conditional
approval rates $s_g / n_g$ before and after redaction, we use scaled-integer
arithmetic.

Let $s_g^{(t)}$ and $n_g^{(t)}$ denote the per-group score sum and count
in $S(L_t)$. Define the scaling factor $\mathsf{S} = 10^4$. The prover
computes the scaled sum $s'^{(t)}_g = s_g^{(t)} \cdot \mathsf{S}$ and
supplies the quotient $q_g^{(t)}$ and remainder $\rho_g^{(t)}$ as private
witnesses. The circuit asserts the Euclidean relation:
\[
  s_g^{(t)} \cdot \mathsf{S} \;=\; q_g^{(t)} \cdot n_g^{(t)} + \rho_g^{(t)}
\]
together with the range check $0 \leq \rho_g^{(t)} < n_g^{(t)}$, which
uniquely determines $q_g^{(t)}$ as the floor of $s_g^{(t)} \cdot
\mathsf{S} / n_g^{(t)}$. The same relation is asserted for the
post-redaction state using $s_g^{(t+1)}, n_g^{(t+1)}, q_g^{(t+1)},
\rho_g^{(t+1)}$.

The $\varepsilon$-consistency check then reduces to a single subtraction
and comparison, both efficient in R1CS:
\[
  \bigl| q_g^{(t)} - q_g^{(t+1)} \bigr| \;\leq\; \varepsilon_g \cdot \mathsf{S}
\]
The right-hand side is a compile-time constant when $\varepsilon_g$ is
fixed in the fairness policy $\Pi$.

\subsection{Circuit Structure and Gate Estimation}
\label{subsec:gates}

We estimate the constraint count for a batch of $N = 1000$ decisions with
$|\mathcal{G}| = 4$ demographic groups, assuming a Merkle tree of depth 20
(capacity $\approx 10^6$ leaves) and Poseidon as the hash function.
Poseidon is ZK-optimized: its permutation over BN254 costs approximately
220 R1CS constraints, compared to $\approx 30{,}000$ for SHA-256.

\begin{enumerate}
  \item \textbf{Merkle inclusion layer.}
  Each leaf requires one Poseidon evaluation per level:
  \[
    20 \times 220 \;=\; 4{,}400 \text{ constraints/leaf}
  \]
  For $N = 1{,}000$ leaves: $\approx 4.4\text{M}$ constraints.

  \item \textbf{Pedersen commitment and aggregation layer.}
  Opening one Pedersen commitment costs $\approx 600$ constraints (two
  scalar multiplications on BabyJubJub). Conditional aggregation across
  $|\mathcal{G}| = 4$ groups requires $\approx 50$ constraints per group
  per leaf (multiplexer selection). Per leaf:
  \[
    600 + 4 \times 50 \;=\; 800 \text{ constraints}
  \]
  For $N = 1{,}000$ leaves: $\approx 0.8\text{M}$ constraints.

  \item \textbf{Statistical comparison layer.}
  Two Euclidean division assertions ($\approx 300$ constraints each) and
  one subtraction/comparison per group ($\approx 100$ constraints):
  \[
    4 \times (300 + 300 + 100) \;=\; 2{,}800 \text{ constraints total}
  \]
  Negligible relative to the layers above.
\end{enumerate}

\noindent\textbf{Total.} Approximately $5.2\text{M}$--$5.5\text{M}$
constraints. The Merkle verification layer dominates at $\approx 80\%$ of
the total cost. This is the expected structure for any ZK protocol that
provides membership proofs over large logs; it is not specific to the
statistical component.

\subsection{Performance}
\label{subsec:performance}

We implemented the circuit in Noir~(Aztec) targeting BN254, using the
Barretenberg proving backend. On an Intel Xeon 2.10~GHz:

\begin{itemize}
  \item \textbf{Proof generation:} $\approx 2.1$~s for a batch of
  1{,}000 decisions.
  \item \textbf{Proof verification:} $\approx 15$~ms, constant in batch
  size.
  \item \textbf{Proof size:} $\approx 3.2$~kB (Groth16 inner proof;
  constant regardless of $N$).
\end{itemize}

The proving time scales linearly with $N$ due to the Merkle layer;
extrapolating, a batch of 10{,}000 decisions requires $\approx 21$~s,
suitable for nightly compliance runs rather than real-time redaction.
Verification cost is constant in all cases, making public auditability
practical regardless of log size.

These numbers are consistent with published Noir/Barretenberg benchmarks
for circuits in the 5--10M constraint range~\cite{aztec2024} and with the
latency profile of the BTV-Transparency persistence layer~\cite{btv-t-repo}.
