% -----------------------------------------------------------------------------
% Section 7: Discussion
% Paper 3 — Accountable Redaction (ZK-SNARKs)
% Security audit patches applied 2026-03-29:
%   Patch 3.1: Attested Demographics proposal (garbage-in mitigation)
%   Patch 3.2: End-to-End Accountability conjecture (formal composition)
% -----------------------------------------------------------------------------

\section{Discussion}
\label{sec:discussion}

\subsection{Limitations}
\label{sec:limitations}

\begin{enumerate}

  \item \textbf{Trusted setup.}
  The Groth16 proving system requires a trusted setup ceremony to generate
  the structured reference string. A dishonest participant could produce a
  backdoored SRS that allows false proofs. We mitigate this by using a
  Powers of Tau ceremony with $\geq 2$ independent participants; a single
  honest participant suffices for soundness.

  \item \textbf{Garbage-in on demographic attributes.}
  The circuit operates on committed demographic attributes. If
  the operator commits false attributes at ingestion time, the
  statistical guarantees are vacuous: a ZK proof of
  $\varepsilon$-consistency over fabricated group labels is
  cryptographically valid but semantically meaningless.
  This is the most significant unmitigated attack vector in the
  current protocol and is structurally analogous to the
  semantic-truthfulness limitation of Paper~1
  (Section~6.1 of~\cite{btv2026}).

  We propose an \emph{Attested Demographics} extension to narrow
  this gap. At ingestion time, the demographic attribute $g_i$ is
  not self-declared by the operator but is accompanied by a digital
  signature $\sigma_{\mathsf{src}}$ from a trusted \emph{attribute
  source}---a civil registry, a government identity provider, or an
  accredited third-party data broker. The ingestion circuit is
  extended to verify:
  \[
    \mathsf{Verify}_{\mathit{pk}_{\mathsf{src}}}(\sigma_{\mathsf{src}},\;
    g_i \,\|\, \mathit{subject\_id}_i) = 1
  \]
  before accepting the Pedersen commitment $C_i = \mathsf{Com}(g_i;\,
  r_i)$. This reduces the trust assumption from ``the operator
  reports demographics honestly'' to ``the attribute source signs
  demographics honestly''---a strictly smaller trust surface, since
  the attribute source has no economic incentive to distort the
  operator's fairness statistics.

  The extension is compatible with the existing circuit: it adds a
  signature verification gate per ingested entry (approximately
  $\sim$25{,}000 constraints for EdDSA, amortised across batch
  ingestion). Implementation and formal analysis of the Attested
  Demographics extension is future work.

  \item \textbf{Proving time.}
  Current Groth16 proving time for the full audit circuit is
  approximately 4.2\,s on an 8-core workstation. Batch proving and
  recursive proof composition (Halo2, Nova) are promising directions
  for reducing this to sub-second latency.

\end{enumerate}

\subsection{Composition with Papers 1 and 2}
\label{sec:composition}

The three papers address accountability at distinct enforcement layers:
Paper~1 at compile-time (structural), Paper~2 at delivery-time
(cryptographic persistence), and this paper at audit-time (statistical
fairness). A system implementing all three layers guarantees: (a)~no decision
without evidence, (b)~no delivery without verifiable persistence, and
(c)~no redaction without statistical consistency.

We conjecture the following composition property, whose formal proof
is future work:

\begin{conjecture}[End-to-End Accountability]
\label{conj:composition}
Let $P$ be a Safe Rust program whose crate graph includes both
\texttt{btv-core} and \texttt{btv-transparency}, operating over a
transparency log $\Sigma$ with redaction governed by the Accountable
Redaction protocol. If $P$ satisfies the Constitutional Enclosure
Theorem (Paper~1), the Persistence Enclosure Theorem (Paper~2), and
$\varepsilon$-Statistical Consistency (this paper), then the
composed system satisfies:
\[
\underbrace{\mathcal{S}(P) = \emptyset}_{\text{no silent decisions}}
\;\land\;
\underbrace{\mathcal{E}(P) = \emptyset}_{\text{no ephemeral verdicts}}
\;\land\;
\underbrace{\forall R,\, g:\; |\Delta\rho_g| \leq \varepsilon_g}_{\text{no selective erasure}}
\]
under the combined trust assumptions of Safe Rust, honest Log Authority
key (Axiom~3.3 of Paper~2), and SNARK knowledge soundness.
\end{conjecture}

\noindent The conjecture is likely to hold because the three predicates
operate at distinct enforcement layers (compile-time, delivery-time,
and audit-time respectively) with non-overlapping trust assumptions.
A formal proof would require showing that no interaction between
layers introduces a new attack surface---for instance, that a
compile-time bypass does not weaken the cryptographic layer.
We leave this as the most important open problem in the series.
