Regulatory frameworks such as EU AI Act Art.~86 and LGPD Art.~18 mandate
verifiable accountability for high-stakes AI decisions. Bound-To-Verdict
(BTV)~\cite{btv2026} demonstrated that \emph{silent decisions}---AI-generated
acts lacking a computationally bound evidence chain---can be eradicated at the
application layer via linear resource types in Safe Rust. However, structural
integrity in memory does not guarantee verifiable durability in the outside
world: an operator who controls the infrastructure can satisfy the BTV invariant
while discarding the decision artifact before it reaches persistent storage,
creating an \emph{Ephemeral Verdict}---a decision legal in its formation but
inaccessible for contestation.

This paper presents \textbf{BTV-Transparency}, a framework that extends the
linear type discipline to the persistence layer. We define the type law
$\mathsf{DeliveryToken} \multimap (V \otimes R)$, enforcing that a decision
cannot be delivered to an end-user without linearly consuming a
\texttt{Verdict}~($V$) and an \texttt{InclusionReceipt}~($R$)---a
cryptographic Ed25519-signed proof of inclusion in a public, append-only log
operated by an independent authority. We prove the \emph{Persistence Enclosure
Theorem}: in any Safe Rust program integrating BTV-Transparency, the set of
Ephemeral Verdicts is empty by construction, across four exhaustive attack
cases (struct-literal forgery, private constructor bypass, silent drop, and
cryptographic MITM). Empirical validation shows that verifiable persistence
adds a mean end-to-end latency of 1.81\,ms (p99:~3.25\,ms)---negligible
relative to the 50--200\,ms inference cost of modern LLM pipelines. Together
with BTV, this work closes the full accountability loop: no decision can exist
without evidence~(Paper~1), and no decision can be delivered without a
verifiable persistence record~(this paper).
