\section{Discussion}
\label{sec:discussion}

\subsection{Scope and Boundaries}

\paragraph{Log key compromise.}
The Persistence Enclosure Theorem holds under Axiom~\ref{ax:log-authority}.
If the Log's private key is compromised, an adversary can forge
\texttt{InclusionReceipt} values. This boundary is deliberately analogous to
the \texttt{unsafe} limitation of BTV~\cite{btv2026}: the type system enforces
structural accountability; cryptographic key security is an orthogonal trust
assumption. In practice, key protection relies on HSM/TPM deployment for the
Log server and independent monitoring (CT Gossip~\cite{rfc6962}) for
misbehaviour detection. Additional mitigations include: (1)~scheduled key
rotation with overlap periods that allow auditors to verify historical receipts;
(2)~requiring submission to $k \geq 2$ independent log authorities (analogous
to the multi-log requirement for TLS certificates in modern browsers), so that
a single key compromise does not unilaterally invalidate the audit trail; and
(3)~CT Gossip-based consistency checks across independent monitors, ensuring
that log misbehaviour or selective suppression is detectable even when the
operator controls the log server infrastructure.

\paragraph{Single-log trust.}
Relying on a single Log authority creates a trust bottleneck. Production
deployments should require submission to $k \geq 2$ independent logs (as CT
browsers require), with the \texttt{DeliveryToken} encoding receipts from all
$k$ logs. This extension preserves the type law; it widens the
\texttt{InclusionReceipt} to a $k$-tuple of receipts and is left as future
work.

\paragraph{Asynchronous submission.}
The current design blocks delivery until the Log responds. For
ultra-low-latency pipelines, a two-phase commit pattern---issue a provisional
token, confirm with the log asynchronously---is possible. The type system can
encode this as a \texttt{PendingDeliveryToken} that must be upgraded to a
\texttt{DeliveryToken} before the payload is released. This extension is
consistent with SCT-based pre-issuance in CT and is left as future work.

\subsection{Relationship to BTV}

BTV-Transparency is a \emph{compositional extension}: it depends on
\texttt{btv-core} as a library and adds a new invariant without modifying or
weakening the Constitutional Enclosure Theorem. Programs that already use BTV
can adopt BTV-Transparency by wrapping their delivery paths in
\texttt{DeliveryToken::seal}. The two theorems compose: a system satisfying
both invariants cannot produce a decision that (a) lacks an evidence chain or
(b) was not verifiably persisted before delivery.

When the transparency log is unavailable, BTV-Transparency fails securely
(Corollary~\ref{cor:fail-secure}): no \texttt{DeliveryToken} can be produced,
so no decision is delivered. In safety-critical deployments (e.g., healthcare)
where hard blocking on log unavailability is unacceptable, the
\texttt{EscalatedVerdict} pattern from BTV~\cite{btv2026} composes naturally
with this fail-secure behaviour: an escalation path can be encoded as a
separate linear type that requires human authorisation, preserving the
accountability guarantee without requiring the log to be reachable.

\subsection{Limitations}

\begin{enumerate}
  \item \textbf{L1 --- In-memory log.} The \texttt{btv-log} reference server is
  not persistent across restarts. Production use requires a durable backend
  (e.g., Trillian + PostgreSQL).
  \item \textbf{L2 --- Loopback latency.} Benchmarks measure loopback
  submission; WAN latency to a remote log will be higher. The system argument
  (log latency is dominated by inference latency) remains valid for remote logs
  in the same cloud region (typical RTT 1--10\,ms). At a representative WAN RTT
  of 10\,ms, the full pipeline would be approximately 15\,ms, representing
  $\leq$30\,\% of a 50\,ms classifier and $\leq$7.5\,\% of a 200\,ms
  generative model---still within acceptable overhead bounds for regulated
  AI applications (see Section~\ref{sec:benchmarks}).
  \item \textbf{L3 --- Single-language boundary.} The enclosure perimeter holds
  within Safe Rust. Polyglot pipelines require the FFI gateway pattern described
  in BTV~\cite{btv2026} Section~6.2.4, extended to wrap the
  \texttt{DeliveryToken} before crossing the language boundary.
  \item \textbf{L4 --- Privacy.} The transparency log stores verdict hashes,
  not verdict content. The SHA-256 fingerprint of the serialised verdict is
  appended to the log; no personal data is transmitted to or stored by
  \texttt{btv-log}. Nonetheless, even hash-only logs may be subject to
  correlation attacks in low-entropy decision spaces. Privacy-preserving
  verification (e.g., zero-knowledge inclusion proofs) is left as future work.
\end{enumerate}
