\section{Implementation}
\label{sec:impl}

This section describes the \texttt{btv-transparency} crate and the lightweight
\texttt{btv-log} server that backs it. Both are written in Safe Rust; the crate
graph contains zero \texttt{unsafe} blocks, verifiable via \texttt{cargo-geiger}.

\subsection{The InclusionReceipt Linear Type}

Listing~\ref{lst:receipt} shows the core types. \texttt{InclusionReceipt} is
linear by construction: it does not implement \texttt{Clone} or \texttt{Copy},
its fields are private (Protection~A), its sole constructor is
\texttt{pub(crate)} (Protection~B), and \texttt{\#[must\_use]} on the struct
prevents silent drops (Protection~C).

\begin{lstlisting}[caption={btv-transparency/src/lib.rs},label={lst:receipt},language=Rust]
use btv_core::Verdict;
use ed25519_dalek::{Signature, VerifyingKey};

#[must_use]
pub struct InclusionReceipt {
    log_index: u64,          // private: Protection A
    signed_tree_head: Signature,
    verifying_key: VerifyingKey,
}

impl InclusionReceipt {
    pub(crate) fn new(index: u64, sth: Signature, key: VerifyingKey) -> Self {
        Self { log_index: index, signed_tree_head: sth, verifying_key: key }
    }

    pub(crate) fn consume(self) -> (u64, Signature) {
        (self.log_index, self.signed_tree_head)
    }
}

#[must_use]
pub struct DeliveryToken {
    verdict: Verdict,
    receipt: InclusionReceipt,
}

impl DeliveryToken {
    pub fn seal(verdict: Verdict, receipt: InclusionReceipt) -> Self {
        Self { verdict, receipt }
    }

    #[must_use = "delivery not confirmed"]
    pub fn deliver(self) -> Result<(Vec<u8>, Vec<u8>), serde_json::Error> {
        let v_json = serde_json::to_vec(&self.verdict)?;
        let r_json = serde_json::to_vec(&self.receipt)?;
        Ok((v_json, r_json))
    }
}
\end{lstlisting}

\subsection{Enclosure Perimeter: Compile-Fail Tests}

The enclosure perimeter is verified by four compile-fail tests in
\texttt{tests/sui/}, mirroring Cases~A--C of the Constitutional Enclosure
Thorem~\cite{btv2026}:

\begin{enumerate}
  \item \texttt{receipt\_struct\_literal.rs} --- struct-literal
  \texttt{InclusionReceipt~\{..\}} rejected with E0451.
  \item \texttt{receipt\_consume\_external.rs} --- external call to
  \texttt{InclusionReceipt::consume} rejected with E0624.
  \item \texttt{receipt\_drop\_silent.rs} --- ignored \texttt{InclusionReceipt}
  promotes \texttt{unused\_must\_use} to hard error.
  \item \texttt{delivery\_without\_receipt.rs} --- constructing
  \texttt{DeliveryToken::seal} without a valid $R$ fails to type-check,
  as no path to \texttt{InclusionReceipt::new} exists outside the crate.
\end{enumerate}

\subsection{The btv-log Server}

\texttt{btv-log} is a lightweight, in-memory Transparency Log server
implementing a minimal CT-style interface:

\begin{itemize}
  \item \texttt{POST /submit} --- accepts a 32-byte verdict hash, appends it to
  the Merkle tree, and returns
  \texttt{\{index, root\_hash, signature\}}.
  \item \texttt{GET /proof/:index} --- returns the Merkle inclusion proof for
  a given index (for auditor verification).
\end{itemize}

The signing key is an Ed25519 keypair generated at startup and configurable
via environment variable for production deployments (HSM injection).
The \texttt{btv-log} server is explicitly scoped as a \emph{reference
implementation}: it stores the Merkle tree in memory and is not designed for
distributed operation. Production deployments should substitute Trillian~\cite{trillian}
or an equivalent distributed log backend.

\subsection{The LogClient}

\texttt{LogClient::submit\_and\_await} sends a synchronous HTTP request to
\texttt{btv-log} and blocks until the \texttt{InclusionReceipt} is returned.
If the log is offline, the call returns \texttt{Err}, and by Theorem~\ref{thm:persistence-enclosure}
no \texttt{DeliveryToken} is produced---the system fails securely, extending
Corollary~4.7 of~\cite{btv2026} to the persistence layer.
