\section{Independence of Powers: Proofs and Counter-Examples}
\label{sec:independence}

We now prove the three independence theorems stated in
Section~\ref{sec:mapping} and contrast each guarantee with the
corresponding failure mode of traditional best-effort architectures.
The counter-examples are not hypothetical: each describes a class of
system that is standard practice in production AI deployments today.

\subsection{Theorem 1: Legislative Independence}
\label{subsec:proof-leg}

\textit{Claim.} The Operator ($\mathcal{E}$) cannot modify the
regulatory constraints encoded in the Constitutional Type System
($\mathcal{L}$) at runtime.

\begin{proof}
Regulatory constraints in BTV are encoded as linear resource types
(e.g., \texttt{EvidenceToken}, \texttt{ComplianceCertificate}) in
the \texttt{btv-core} crate~\cite{btv2026}. By Rust's type system
semantics, type definitions are resolved at compile time and baked
into the binary artifact; they admit no runtime reflection or
modification. We trace the operator's capability:

\begin{enumerate}
  \item Suppose $\mathcal{E}$ attempts to violate Legislative
  Independence by altering the constraint $V \multimap E \otimes C$
  so that a \texttt{Verdict} may be constructed without consuming an
  \texttt{EvidenceToken}.
  \item This requires modifying the type definition in
  \texttt{btv-core/src/types.rs} and recompiling the crate.
  \item Recompilation produces a binary with a distinct
  cryptographic hash $h'\neq h$, where $h$ is the hash of the
  constitutionally-verified binary published to $\Sigma$ at
  deployment~\cite{btv-t-repo}.
  \item The Transparency Log $\Sigma$ contains the deployment
  entry $\langle h, t_0, \sigma_{\mathcal{L}} \rangle$, where
  $\sigma_{\mathcal{L}}$ is a signature over both the binary hash
  and the type context $\Gamma$.
  \item Any auditor $\mathcal{J}$ observing $h' \neq h$ in the
  log detects the substitution. By Theorem~\ref{thm:jud}, this
  detection requires no cooperation from $\mathcal{E}$.
\end{enumerate}

Therefore, legislative modification requires a public, detectable
``constitutional amendment'' (a new log entry with a new binary hash),
not a silent runtime change. $P_{\mathcal{E}} \cap P_{\mathcal{L}}
= \emptyset$ at runtime. \qed
\end{proof}

\paragraph{Counter-example (status quo).}
In a standard Python-based AI pipeline, compliance rules are
implemented as policy configuration files (\texttt{config.json}) or
branch predicates (\texttt{if threshold > 0.7}). The operator, having
write access to the production environment, can modify
\texttt{config.json} in-place without recompilation, without a new
deployment event, and without any auditable record of the change.
The ``Legislature'' (the configuration file) resides within the
``Executive'' jurisdiction (the production server). Legislative
Independence is structurally impossible: the same principal who executes
decisions also controls the rules that govern them.

\subsection{Theorem 2: Judicial Independence}
\label{subsec:proof-jud}

\textit{Claim.} An Auditor ($\mathcal{J}$) can verify the integrity of
any state transition $\sigma \in \Sigma$ without permission,
cooperation, or data access from the Operator ($\mathcal{E}$), and
without learning the private attributes of any decision subject.

\begin{proof}
Judicial Independence decomposes into two sub-claims: access
independence and content verifiability.

\textbf{(a) Access independence.}
The Transparency Log $\Sigma$ is maintained as a public
content-addressed Merkle tree~\cite{btv-t-repo}. The root
$r_t$ is published to a public bulletin at each epoch $t$.
To verify that transition $\sigma_i$ is included in $\Sigma$,
the auditor constructs a Merkle inclusion proof using only
$r_t$ and the public components of $\sigma_i$. The verification
algorithm is:
\[
  \mathsf{MerkleVerify}(r_t,\; \pi^{\mathsf{path}}_i,\;
  H(\sigma_i)) \in \{0,1\}
\]
This computation requires no API call to $\mathcal{E}$, no
authorisation token, and no credential controlled by the operator.
It runs on the auditor's local hardware against public data.

\textbf{(b) Content verifiability (statistical compliance).}
To verify statistical non-discrimination across protected attributes
(e.g., that approval rates are race-neutral), the auditor
obtains a ZK batch proof $\pi^{\mathsf{stat}}$ from the operator
(Paper~3~\cite{btv-ar2026}):
\[
  \mathsf{ZKVerify}\!\left(
    \pi^{\mathsf{stat}},\;
    \mathsf{stmt} = \left[\,
      \frac{|\{\sigma_i : d_i = 1, a_i = 0\}|}
           {|\{\sigma_i : a_i = 0\}|} < \varepsilon
    \,\right]
  \right) = 1
\]
where $a_i$ is a protected attribute and $d_i$ the decision.
By the soundness of the ZK proof system~\cite{btv-ar2026}, a
successful verification convinces the auditor of the statistical
statement without revealing any individual's value of $a_i$ or $d_i$.
The proof is provided by the operator but its validity is
operator-independent: a forged proof for a false statement is
indistinguishable from a valid proof only with probability negligible
in the security parameter $\lambda$.

Combining (a) and (b): $\mathcal{J}$ verifies both structural inclusion
and statistical compliance without querying $\mathcal{E}$ and without
learning private data. $P_{\mathcal{J}} \cap P_{\mathcal{E}} =
\emptyset$. \qed
\end{proof}

\paragraph{Counter-example (status quo).}
In a standard enterprise SQL database, the auditor must request
\texttt{SELECT} access from the database administrator, who reports to
the operator. The operator may legitimately deny access citing trade
secret protection or GDPR data minimisation requirements. Even when
access is granted, the operator can create filtered views that exclude
incriminating rows, set \texttt{row\_security} policies that silently
redact records, or purge logs before the audit window. The judiciary
is constitutionally subservient to the executive's permission to see.
This is the configuration that produced the evidence failures catalogued
in Section~\ref{sec:forensic} of Paper~4~\cite{btv-econ}:
in 9 of 20 enforcement cases, the auditor's inability to access records
was the primary obstacle to timely resolution.

\subsection{Theorem 3: Executive Containment}
\label{subsec:proof-exec}

\textit{Claim.} The Operator ($\mathcal{E}$) cannot execute a
consequential decision $d$ without generating a non-repudiable
evidence receipt committed to $\Sigma$.

\begin{proof}
We trace the only valid code path from decision computation to
external delivery~\cite{btv2026}:

\begin{enumerate}
  \item \texttt{DeliveryToken} is a linear type authorising the
  external API response. By linearity, it must be consumed exactly
  once; it cannot be duplicated or discarded.
  \item \texttt{DeliveryToken::seal()} requires consuming an
  \texttt{InclusionReceipt} as an argument.
  \item \texttt{InclusionReceipt} is produced exclusively by the
  Log Server's \texttt{append()} method, which returns a receipt only
  after the entry is committed to $\Sigma$ and the Merkle root is
  updated.
  \item Therefore, the execution path
  $\mathtt{Decide} \to \mathtt{Deliver}$ is type-constrained to pass
  through $\mathtt{Log} \to \mathtt{Receipt}$:
  \[
    \mathtt{decide}() \to V \multimap E \otimes C
    \xrightarrow{\mathtt{append}} \mathtt{InclusionReceipt}
    \xrightarrow{\mathtt{seal}} \mathtt{DeliveryToken}
    \xrightarrow{\mathtt{send}} \bot
  \]
  \item Any attempt to short-circuit this path---constructing a
  \texttt{DeliveryToken} without an \texttt{InclusionReceipt}---is a
  compile-time type error (Rust errors E0308, E0061). The
  executable binary cannot contain this path.
\end{enumerate}

Since no valid binary can deliver a decision without a committed
receipt, and since any binary substitution is detectable by
Theorem~\ref{thm:jud}, executive action without logging is not merely
prohibited---it is architecturally unreachable. \qed
\end{proof}

\paragraph{Counter-example (status quo).}
In standard logging practice, the developer writes
\texttt{logger.info(``Decision: \{result\}'')} \emph{after} the
decision logic and \emph{before} the API response. This logging
call is a procedural instruction, not a structural constraint:
under high load, the logging buffer may overflow and silently drop
entries; under infrastructure failure, the log service may be
unavailable while the API continues to serve decisions; under
malicious intent, a developer can comment out the line or add a
conditional that suppresses logging for a specific cohort. In
all three scenarios, the executive acts without leaving a
fooprint, and no compile-time or runtime mechanism prevents it.
The Amazon hiring case (Paper~4, \S2.2~\cite{btv-econ}) is an
instance of this failure: the bias mechanism operated for
approximately four years before a procedural audit---not a structural
one---detected it.

\subsection{Corollary: Constitutional Completeness}
\label{subsec:completeness-proof}

\begin{corollary}[Constitutional Completeness, restated]
In a BTV institution $\mathcal{I}$, no single principal holds
capabilities in more than one constitutional branch simultaneously:
\[
  P_{\mathcal{L}} \cap P_{\mathcal{E}} \cap P_{\mathcal{J}} = \emptyset
\]
The Operator may \emph{act} but not \emph{conceal} or
\emph{redefine rules}. The Regulator may \emph{define rules} but not
\emph{act} or \emph{audit}. The Auditor may \emph{verify} but not
\emph{act} or \emph{alter rules}.
\end{corollary}

\begin{proof}
Direct from Theorems~\ref{thm:leg}, \ref{thm:jud}, \ref{thm:exec}.
The pairwise disjointness of all three capability sets follows from the
independence proofs above. The $\mathcal{J} \cap \mathcal{L}$
disjointness follows by construction: the auditor's verification
function $V : \Sigma \to \{0,1\}$ is read-only over $\Sigma$ and has
no write capability over $\mathcal{L}$; this is a capability that the
auditor never requests and the type system never grants.
\end{proof}

\begin{remark}[Novelty of the Enforcement Mechanism]
Montesquieu's original formulation relies on \emph{institutional
design} and \emph{social contract} to enforce separation: no formal
mechanism prevents a judge from also acting as legislator, beyond
political norms and constitutional law. Constitutional Code is the
first realisation in which the separation is enforced by the
mathematical properties of the execution substrate---specifically,
linear type theory (Theorems~\ref{thm:leg},~\ref{thm:exec}) and
cryptographic soundness (Theorem~\ref{thm:jud}). The institution's
constitution cannot be violated because the violation is not a
programme that can be compiled and run.
\end{remark}
