\section{Implications for Regulators}
\label{sec:regulators}

The Constitutional Code framework does not merely offer a technical
solution; it mandates a paradigm shift in regulatory practice. Under
traditional governance models, regulators act as \emph{post-hoc}
auditors, arriving after the harm has occurred and attempting to
reconstruct decisions from incomplete records---the configuration that
produced the 20 enforcement failures analysed in Paper~4~\cite{btv-econ}.
Under Constitutional Code, regulators become \emph{ex-ante} architects,
defining the type invariants that constrain the system before it
executes a single decision. This section formalises that role and
derives three operational consequences.

\subsection{From Policy Documents to Type Definitions}
\label{subsec:types-as-law}

Current AI regulations are written in natural language. The EU AI
Act~\cite{eu-ai-act} Articles~6--7 require ``transparency'' and
``accuracy''; the LGPD~\cite{lgpd} Article~20 mandates ``clear
information'' about automated decisions. These requirements are
necessary but structurally unenforceable: they are satisfied by
compliance reports, not by code, and therefore depend on the
operator's good faith---precisely the assumption that the status quo
counter-examples of Section~\ref{sec:independence} show to be
insufficient.

In the Constitutional Code framework, each statutory requirement
corresponds to a type-level invariant:

\begin{table}[h]
\centering
\caption{Statutory Requirements as Type Invariants}
\label{tab:law-to-types}
\begin{tabular}{p{3.8cm} p{4.2cm} p{2.6cm}}
\toprule
\textbf{Statutory Requirement} & \textbf{Constitutional Type} & \textbf{Source} \\
\midrule
``Record the decision'' & \texttt{Verdict $\multimap$ E $\otimes$ C} & EU AI Act Art.~12 \\
``Provide a reason'' & \texttt{EvidenceToken} schema & LGPD Art.~20 \\
``Enable human override'' & \texttt{OperatorOverrideToken} & EU AI Act Art.~14 \\
``Preserve the log'' & Append-only Merkle commitment & GDPR Art.~5(e) \\
``Allow audit'' & Public root + ZK batch proof & EU AI Act Art.~67 \\
\bottomrule
\end{tabular}
\end{table}

The regulatory burden shifts from verifying \emph{compliance reports}
(a periodic, resource-intensive process) to verifying
\emph{interface definitions} (a one-time type-checking operation).
A regulator does not need to parse millions of lines of source code;
they need only verify that the declared types in the operator's
published \texttt{btv-core} dependency satisfy the statutory schema.
Compliance becomes a property of the compilation artefact, not a
property of the operator's intentions.

\subsection{The Constituent Role}
\label{subsec:constituent}

We formalise the regulator's function under Constitutional Code as the
\emph{Constituent}: the actor who defines the constitutional rules
before the institution operates, analogous to a constitutional
convention that precedes the legislature, executive, and judiciary
it creates.

\begin{enumerate}
  \item \textbf{Definition (Constitutional Types).} The regulator
  publishes a \emph{Regulatory Type Schema} $\mathcal{L}_{\mathsf{reg}}$:
  a set of linear type constraints that any compliant BTV deployment
  must satisfy. This is analogous to publishing building codes before
  issuing a construction permit.

  \item \textbf{Certification (Binary Attestation).} The regulator
  certifies the hash $h$ of the compiled operator binary as
  ``Constitutionally Compliant'' after verifying that the binary's
  type context $\Gamma$ satisfies $\mathcal{L}_{\mathsf{reg}}$. This
  certification is published to a public registry. Operating an
  uncertified binary becomes a strict liability offense---the
  algorithmic equivalent of a building without a permit.

  \item \textbf{Monitoring (Judicial Accreditation).} The regulator
  accredits independent Monitor organisations to serve as the
  Judicial branch: they subscribe to the operator's public Merkle
  root, run continuous ZK verification of batch proofs, and report
  deviations. The regulator does not perform monitoring itself;
  it defines the qualification criteria for Monitors and maintains
  their registry.
\end{enumerate}

This architecture resolves the \emph{resource asymmetry problem} that
undermines conventional AI oversight: regulatory agencies are
systematically understaffed relative to the organisations they supervise.
By enforcing structural constraints (types) rather than process
constraints (audits), the regulator uses the compiler as a force
multiplier. A single type schema, once published, enforces 100\%
coverage of the defined rules across all deployments of the certified
binary---without requiring the regulator to inspect individual decisions.

\subsection{The Compliance Ladder}
\label{subsec:ladder}

Section~\ref{sec:credit} of Paper~4~\cite{btv-econ} introduced the
Compliance Credit $\delta$ as an economic mechanism to internalise
the externalities of opacity. Constitutional Code provides the
mechanism to \emph{standardise} that credit across regulatory regimes.

We define three tiers of constitutional compliance:

\begin{table}[h]
\centering
\caption{Constitutional Compliance Tiers and Credit Levels}
\label{tab:ladder}
\begin{tabular}{clp{4.8cm}c}
\toprule
\textbf{Level} & \textbf{Label} & \textbf{Structural Guarantee} & \textbf{Credit $\delta$} \\
\midrule
0 & Opaque & No structural accountability. Fine exposure unmitigated. & $0$ \\
1 & Accountable & Papers~1--2: Linear types + append-only log. Evidence cannot be suppressed. & $0.4$ \\
2 & Constitutional & Papers~1--3: Full stack with ZK audit proofs. Judicial branch fully independent. & $0.9$ \\
\bottomrule
\end{tabular}
\end{table}

The compliance ladder creates a \emph{regulatory gradient}: operators
are financially incentivised to ascend toward higher verifiability
levels, because $\delta$ directly reduces expected fine liability
(Paper~4, Theorem~5.1~\cite{btv-econ}). Compliance transforms from a
cost centre into a risk-mitigation asset with a computable expected
return on investment.

Critically, the $\delta$ credits are not granted on the basis of
operator self-certification---they are granted on the basis of binary
attestation (\S\ref{subsec:constituent}). An operator claiming Level~2
but running an uncertified binary is detectable by any Monitor via
hash comparison. The credit system is structurally fraud-resistant.

\subsection{The Dissolution of the Trade Secret Defence}
\label{subsec:trade-secret}

The most persistent structural barrier to AI accountability litigation
is the \emph{trade secret defence}: operators assert that their
algorithm constitutes proprietary intellectual property and that
disclosure to auditors would cause competitive harm~\cite{pasquale2015}.
Courts in both the EU and US have partially accepted this argument,
blocking full algorithm disclosure in several enforcement actions.

Constitutional Code dissolves this tension structurally. Because the
Judicial branch verifies the system via Zero-Knowledge
proofs~\cite{btv-ar2026}, the auditor confirms the \emph{statistical
behaviour} and \emph{structural compliance} of the algorithm without
ever accessing the source code, the model weights, or the training data.
Formally:
\[
  \mathsf{ZKVerify}(\pi^{\mathsf{stat}},\; \mathsf{stmt}) = 1
  \quad \Rightarrow \quad
  \text{Auditor learns: stmt is true; learns nothing else.}
\]
The operator retains full intellectual property protection over the
algorithm's implementation. Society gains a cryptographic guarantee
of the algorithm's fairness properties. The trade secret defence
becomes obsolete as a shield against accountability---not because
the law changed, but because the architecture renders disclosure
unnecessary for verification.

This represents a Pareto improvement over the status quo: the operator
is not worse off (IP is protected), the regulator is not worse off
(compliance is verifiable), and decision subjects are strictly better
off (fairness is guaranteed, not merely asserted).
