% -----------------------------------------------------------------------------
% Section 6: Discussion
% Paper: "Silent Decisions Are Type Errors: Enforcing AI Accountability
%         via Linear Resource Types"
% Venue: IEEE Computer — Special Issue on AI Governance
% Deadline: 13 April 2026
% Authors: Daniel Lau Pereira Soares
% Rev 2: 24 March 2026 — four reviewer-hardening revisions:
%   Gap 1: AttestableContext trait for semantic truthfulness
%   Gap 2: FFI gateway + BTV sidecar for polyglot pipelines
%   Gap 3: ComplianceAuthority factory replaces public constructor
%   Gap 4: CAL demoted to design trade-off; CAP analogy removed
% Rev 3: 29 March 2026 — security audit patches applied:
%   Patch 1.1: Trust-assumption alignment with Paper 2 (BTV_HMAC_KEY)
%   Patch 1.2: ComplianceAuthority factory (implemented, not proposed)
%   Patch 1.3: Sixth future-work item — formal composition theorem
% -----------------------------------------------------------------------------

\section{Discussion}
\label{sec:discussion}

The Constitutional Enclosure Theorem establishes a strong structural
guarantee within a precisely bounded scope.
This section examines the boundaries of that scope honestly,
identifies limitations of the current reference implementation,
and proposes concrete extensions that preserve the core invariant
while closing each gap.

% ----------------------------------------------------------------
\subsection{Scope and Boundaries of the Guarantee}
\label{sec:scope}

\paragraph{Unsafe Rust.}
The \texttt{unsafe} keyword grants the programmer the ability to
bypass borrow-checker and visibility rules, including the private
field restrictions that enforce the encapsulation boundary
(Section~\ref{sec:linear_types}).
Production deployments should enforce a zero-\texttt{unsafe} policy
over the transitive dependency graph, auditable via
\texttt{cargo geiger}~\cite{cargogeiger}.
This is an operational policy decision, not a theoretical barrier:
the tooling exists and its enforcement requires no changes to the
BTV crate.

\paragraph{Semantic truthfulness and the AttestableContext extension.}
The theorem guarantees that \emph{some} byte sequence was hashed
and bound to a verdict---not that the sequence faithfully represents
the context presented to the AI system at inference time.
A caller could pass \texttt{b"placeholder"} to
\texttt{EvidenceToken::new} and obtain a structurally valid token.

We propose a concrete extension that narrows this gap without
violating the core invariant.
Introduce a sealed \texttt{AttestableContext} trait:

\begin{lstlisting}[language=Rust,caption={AttestableContext trait
  (proposed extension)}]
pub trait AttestableContext: Sealed {
    fn canonical_bytes(&self) -> &[u8];
    fn schema_version(&self)  -> u16;
    fn source_attestation(&self) -> SourceAttestation;
}
\end{lstlisting}

\noindent
where \texttt{SourceAttestation} encapsulates a digital signature
produced by the inference service over the payload, using a key
stored in an HSM or TPM.
\texttt{EvidenceToken::new} would then accept
\texttt{impl AttestableContext} instead of \texttt{\&[u8]}.
Forging evidence would require compromising the signing key---a
problem reducible to PKI security---rather than simply passing
an arbitrary byte string.

This extension tightens the type law to
$V \multimap (E_{\text{attested}} \otimes C)$,
which is strictly stronger than $V \multimap (E \otimes C)$:
every proof of the original invariant is also a proof of the
attested variant, but not vice versa.
The \texttt{Sealed} supertrait prevents external crates from
implementing \texttt{AttestableContext}, preserving the closed
perimeter of Axiom~3.

% ----------------------------------------------------------------
\subsection{Limitations of the Reference Implementation}
\label{sec:limitations}

The \texttt{silent-decisions-proof} kernel is a proof-of-concept
designed to validate the theoretical claims of this paper.
Four implementation choices trade production robustness for
clarity of exposition; each has a well-defined remediation path.

\paragraph{L1 — Static HMAC key (mitigated).}
The reference implementation now reads the HMAC key from the
\texttt{BTV\_HMAC\_KEY} environment variable at runtime, falling back
to a compile-time constant only when the variable is unset.
This closes L1 for deployments that inject a key from an HSM or KMS.
However, L1 also surfaces an architectural asymmetry with the trust
model of BTV-Transparency~\cite{btv-t-repo}: Paper~2's Persistence
Enclosure Theorem relies on Axiom~3.3 (Log Authority), which assumes
the Log's Ed25519 private key is held in an HSM inaccessible to the
System Operator. The verdict integrity seal offers the same
guarantee only when \texttt{BTV\_HMAC\_KEY} is provisioned from the
same HSM infrastructure that protects the Log Authority key.
Without this alignment, the system's overall non-repudiation
guarantee is bounded by the weakest key-management practice.
The type-level guarantee (Theorem~\ref{thm:enclosure}) is independent
of key management and remains valid regardless of the HMAC key source.

\paragraph{L2 — \texttt{ComplianceAuthority} factory (implemented).}
The \texttt{ComplianceToken::new} constructor was originally
\texttt{pub}, allowing any external crate to self-declare a
jurisdiction and policy version without validation.
The reference implementation now restricts this constructor to
\texttt{pub(crate)} and exposes a \texttt{ComplianceAuthority}
factory that validates jurisdiction strings against an allowlist
of recognised regulatory regimes (\texttt{BR-LGPD},
\texttt{EU-GDPR}, \texttt{EU-AI-ACT}).
The authority's signing key is read from the
\texttt{BTV\_AUTHORITY\_KEY} environment variable, with a
deterministic fallback for the proof test suite.
Two additional proof clauses (Clauses~16 and~17) verify that
(a)~an unrecognised jurisdiction (\texttt{"Narnia"}) is rejected,
and (b)~valid jurisdictions produce well-formed tokens.
The type law tightens to
$V \multimap (E \otimes C_{\text{signed}})$; the encapsulation
boundary (Axioms~1--3) is preserved.

\paragraph{L3 — Single-node evidence chain.}
The current \texttt{EvidenceToken} captures a single BLAKE3 hash
of the decision context.
Multi-step AI pipelines (retrieval, re-ranking, generation,
scoring) produce intermediate states that each carry accountability
implications.
Extending the type system to a \emph{chain} of evidence tokens
where each pipeline stage consumes the prior token and produces
a new one is a natural generalisation that preserves
$V \multimap (E \otimes C)$ at each step.

\paragraph{L4 — Polyglot pipelines.}
Production AI systems rarely execute in a single-language process.
Python orchestrators, C++ inference engines, and Java API gateways
must all participate in the same governance boundary.
Because the theorem holds within Safe Rust (Axiom 3), external
code constitutes $\Gamma_{\text{ext}}$ by definition, and cannot
construct a \texttt{Verdict} directly.

We propose two complementary patterns.
First, an \emph{accountability enclave}: a Rust gateway crate
exposing a C-ABI boundary,

\begin{lstlisting}[language=Rust,caption={FFI gateway
  (accountability enclave pattern)}]
#[no_mangle]
pub extern "C" fn btv_create_verdict(
    context_ptr:  *const u8,
    context_len:  usize,
    jurisdiction: *const c_char,
    // ...
) -> *mut BtvVerdictHandle { ... }
\end{lstlisting}

\noindent
External code in any language requests a verdict from the
gateway; the gateway applies all BTV invariants internally.
This is structurally analogous to how SQLite enforces ACID
properties regardless of the calling language: the
guarantees are encoded inside the engine, not in the caller.

Second, a \emph{BTV sidecar} deployed alongside each service in a
multi-service gRPC pipeline (analogous to Envoy/Istio mTLS
proxies), which intercepts decision calls and enforces the
evidence-chain protocol at each hop.
This pattern addresses L3 and L4 jointly: the sidecar
constructs a new \texttt{EvidenceToken} at each pipeline stage,
chaining them into the multi-step evidence structure.
In both patterns, the Enclosure Theorem is preserved: external
callers remain in $\Gamma_{\text{ext}}$ and cannot bypass
\texttt{Verdict::new}.

% ----------------------------------------------------------------
\subsection{CAL Design Trade-Offs}
\label{sec:cal}

The four limitations above reflect a deliberate set of design
choices made in the reference implementation.
We observe that a production governance kernel faces tensions among
three desirable properties:
\textbf{C}onsistency (every materialized verdict has a causally-bound
evidence chain),
\textbf{A}vailability (the system can always issue a verdict in
finite time), and
\textbf{L}egality (tokens are issued only by a verified authority).
In the reference implementation, Consistency and Legality are
prioritised: when evidence cannot be produced or the compliance
authority is unreachable, the system fails securely rather than
issuing an unevidenced verdict, at the cost of reduced availability.

We present this as a practical design observation rather than a
formal impossibility result.
Whether a result analogous to the FLP impossibility theorem~\cite{flp1985}
can be proved for accountability systems under asynchronous network
conditions is an open question we identify as future work.

To partially address the availability cost, the reference
implementation introduces \texttt{EscalatedVerdict}---a sibling type
to \texttt{Verdict} that formalises human escalation as a
linearly-typed alternative.
When an \texttt{EvidenceToken} cannot be produced (e.g., model
timeout, unavailable context), the developer may construct an
\texttt{EscalatedVerdict} by consuming exactly one
\texttt{OperatorToken}: a cryptographic attestation binding a named
human operator to the decision act.
The type law for the escalated path is
$V_{\mathrm{esc}} \multimap (O \otimes \mathbf{1})$,
where $O$ is the sole linear resource and $\mathbf{1}$ denotes the
ordinary (non-linear) parameters (decision, context reference, reason).
Corollary~\ref{cor:escalated} (Section~\ref{sec:escalated}) proves
that this path preserves $\mathcal{S} = \emptyset$: no
\texttt{EscalatedVerdict} can be constructed without consuming its
\texttt{OperatorToken}, and the four compile-fail tests mirror
Cases~A--C of the main theorem.
The measured overhead is comparable:
\texttt{size\_of::\allowbreak<EscalatedVerdict>()} $= 160$~bytes
versus $152$~bytes for \texttt{Verdict}, confirming that structured
human oversight introduces negligible memory cost.
This extension structurally aligns with EU AI Act Art.~14 (human
oversight for high-risk systems)~\cite{eu2024aiact}: the system
cannot silently fall back to an unevidenced automated verdict---it
must either produce full automated evidence or explicitly escalate
to a named, cryptographically bound human operator.

% ----------------------------------------------------------------
\subsection{Future Work}
\label{sec:future}

Five directions are most immediate.
First, implementing the \texttt{AttestableContext} extension
(Section~\ref{sec:scope}) and the \texttt{ComplianceAuthority}
factory (L2) would close the semantic-truthfulness and
self-declaration gaps in the reference kernel.
Second, extending the evidence chain to multi-step pipelines via
the sidecar pattern (L3, L4) would make the framework applicable
to retrieval-augmented generation and agentic AI architectures.
Third, formalising the type-system extension in a proof assistant
(Lean~4 or Coq) would elevate the Constitutional Enclosure
Theorem from a machine-checked Rust proof to a fully mechanised
mathematical proof.
Fourth, investigating whether a formal CAL impossibility result
holds under asynchronous network conditions would provide
theoretical grounding for the design trade-offs observed in
Section~\ref{sec:cal}.
Fifth, unifying \texttt{Verdict} and \texttt{EscalatedVerdict}
under the \texttt{AccountableDecision} trait and formalising the
composed invariant $(V_{\mathrm{auto}} \oplus V_{\mathrm{esc}})
\multimap R$, where $R$ is the linear resource appropriate to
each branch, would complete the type-theoretic account of
mixed human-automated decision pipelines.
Sixth, proving a formal composition theorem showing that the
Constitutional Enclosure (this paper), Persistence Enclosure
(Paper~2~\cite{btv-t-repo}), and $\varepsilon$-Statistical
Consistency (Paper~3~\cite{btv-ar2026}) compose into a single
end-to-end accountability invariant---specifically, that a system
satisfying all three predicates simultaneously guarantees
$\mathcal{S} = \emptyset \;\land\; \mathcal{E} = \emptyset
\;\land\; \forall R:\,|\Delta\rho_g| \leq \varepsilon_g$---would
close the composition gap identified in Paper~3
(Section~7.3 of~\cite{btv-ar2026}) and provide the unified
security statement that the series currently lacks.
